Functional Pearl: Parametricity in the Presence of GADTs

### 2021-03-08 20-gadts/GADT-story.tex
final version of paper?

\title[Parametricity in the Presence of GADTs]{Parametricity in the
  Presence of GADTs}

@@

@@

{\color{blue} Combine these two paragraphs.}

We have shown that GADTs can be considered as data types in two
contrasting ways: as completely determined by their syntax, in which
case they are represented by their Church encodings, or as the
functorial completion of their syntax, in which case they are
represented as primitive data types.  But, regardless of which
representation we choose, GADTs fail to be have the expected
parametricity properties. Indeed, we can read the results of
Section~\ref{sec:par} as showing that i) some standard parametricity
results (e.g., inhabitation results) for calculi representing GADTs as
their Church encodings cannot be extended to calculi that represent
GADTs as primitive data types may not be possible, and ii) some
naturality results (e.g., rearrange-transform properties) for calculi
representing GADTs as primitive cannot be extended to calculi that
represent GADTs as their Church encodings. So, with neither
representation can we obtain {\em both} naturality properties {\em
  and} non-naturality consequences of parametricity. Intuitively
speaking, this is because a GADT is not a data type in the usual sense
of being a container, and completing it so that it becomes one
destroys the relationally uniform behavior it must exhibit in a
parametric model. Which representation of GADTs the user chooses can
therefore have profound consequences for how they are able to program
with, and reason about, them.

All is not lost, however. Despite there being no parametric models for
calculi incorporating GADTs as primitive data types, we still get all
possible rearrange-transform properties for these types. This means
that analogues of most of Wadler's free theorems for ADTs hold for
GADTs, since almost all of the specific consequences of parametricity
given in~\cite{wad89} (see especially Figure~1) are actually of this
form for list types. In fact, even though such naturality properties
are generally regarded as consequences of {\em parametricity}, they
are actually consequences of data types' {\em functorial semantics}.
On the other hand, despite there being no models interpreting type
constructors as functors for calculi incorporating GADTs as Church
encodings, and thus there being no models validating all
rearrange-transform properties for these types, parametric models for
such calculi do exist, and these can be used, e.g., to determine which
instances of a GADT are inhabited.

{\color{blue} Comparison with Stephanie and Dmitrios? Comparison with
  Bob's Parametric Model? Other conclusions????}

>>>>

\title[Functional Pearl: Parametricity in the Presence of
  GADTs]{Functional Pearl: Parametricity in the Presence of GADTs}

@@

\vspace*{-0.05in}

@@

We have shown that GADTs can be considered as data types in two
contrasting ways: as completely determined by their syntax, in which
case they are represented by their Church encodings, or as the
functorial completion of their syntax, in which case they are
represented as primitive data types.  Significantly, regardless of
which representation we choose, GADTs fail to have {\em all} expected
parametricity properties. Indeed, we can read the results of
Section~\ref{sec:par} as showing that i)
%some standard parametricity results (e.g.,
inhabitation results for calculi representing GADTs as Church
encodings cannot be extended to calculi representing them as primitive
data types, and ii)
%some
naturality results
%(e.g., rearrange-transform properties)
for calculi representing GADTs as primitive data types cannot be
extended to calculi representing them as Church encodings. So, with
neither representation can we obtain {\em both} naturality
consequences {\em and} non-naturality consequences of
parametricity. Intuitively, this is because a GADT is not a data type
in the usual ``container'' sense, and completing it so that it becomes
one destroys its relationally uniform --- i.e., parametric ---
behavior.
%required in a parametric model for parametricity.
Which representation of GADTs a user chooses can therefore have
profound consequences for programming with, and reasoning about, them.

Although there are no parametric models for calculi representing GADTs
as primitive data types, GADT analogues of most of Wadler's free
theorems for ADTs still hold.
%all possible rearrange-transform properties
%in such calculi since
%This means that analogues of most of Wadler's free theorems for ADTs
%hold for GADTs,
Indeed, most of the parametricity results in Figure~1 of~\cite{wad89}
are actually naturality properties.
%of this form for list types. In fact, even though such
Naturality properties are often regarded as consequences of {\em
  parametricity}, but are, in fact, derivable directly from data
types' {\em functorial semantics}. Even for data types whose Church
encodings and primitive representations coincide, the analysis in this
paper clearly distinguishes those free theorems that are consequences
of naturality from those that are consequences of parametricity {\em
  per se}.

\begin{comment}
On the other hand, despite there
being no models interpreting type constructors as functors for calculi
incorporating GADTs as Church encodings,
%and thus there being no models validating all
%rearrange-transform properties for these types,
parametric models for such calculi do exist, and these can be used,
e.g., to determine which instances of a GADT are inhabited.
\end{comment}

Atkey's parametric model for $F_\omega$ represents data types as
Church encodings. It requires the user to supply a $\mathsf{map}$
function for the type constructor whose fixpoint characterizes the
data type. But, importantly, functoriality of an underlying type
constructor does not imply functoriality of its fixpoint, so the data
type itself still need not necessarily support a
$\mathsf{map}$. Similarly,~\cite{vw10} presents a parametric model for
an extension of $F_\omega$ that supports type equality, and can encode
GADTs. Functoriality is not guaranteed here, either; in fact, the
parametric properties of GADTs described in the preliminary
work~\cite{vw06} are all inhabitation results, rather than naturality
results. Finally,~\cite{ms09} encodes GADTs as Scott encodings rather
than Church encodings but, again, cites only inhabitation results for
them.
